---
title: "R Notebook"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---
```{r include=FALSE}
library(tidyverse)
library(lubridate)
library(move)
library(momentuHMM)
library(leaflet)
library(RColorBrewer)
library(sf)
library(furrr)
library(here)
library(crawl)
```

Load data for use in the assessment
```{r}
load(file=here("classifier_fall", "fall_all_data.RData"))
```

Determine which individuals *didn't* die too early (here defined as 12/15, probably need to bump this up to 1/15)
```{r}
id_list <- unique(amwo_hmm$ID)
id_list <- id_list[!(id_list %in% individuals_died_too_early)]
```

Load the results object
```{r}
results <- readRDS(here("classifier_fall", "fall_all_4_state_model.rds"))
```

Frontmatter: prepare the model

Label states and create distributions for movements
```{r}
stateNames <- c("pre","migration", "stopover", "post")
dist <- list(step="gamma", angle="wrpcauchy")
```

fix step length parameters and state transitions
```{r}
fixPar <- list(beta=matrix(c(NA, -1000, -1000, # pre-migration to transitions, can't go pre- to post
                             -1000, NA, NA,  #once entered migration can't go back to pre
                             -1000, NA, -1000, #Once entered stopover, can only go back to migration
                             -1000, -1000, -1000), nrow = 1, byrow = TRUE)) #once entered post-migration, can't leave

```

These serve as the starting parameters for the HMM to optimize.
```{r}
Par0_m1 <- list(step=c(0.243451270, 3.008102e+02, 0.952588671, 0.322758191, #mean in km
                       0.371507995, 2.812091e+02, 1.722086734, 0.527213629, # SD in km
                       0.004520002, 3.196105e-67, 0.002780489, 0.002684457), #zeromass ie how many zeroes expected in the distribution
                angle=c(-0.0002303861, -0.0930466, 0.01802796, 0.0004041137, #mean
                        0.9251369010, 0.3917246, 0.18833891, 0.7549546329)) #concentration
```

Create a function that fits the model, predicts states for each step (predicted and observed) and input in the dataset. Note: the listed x y is the starting location of the step. Then use this to determine what the bird's state was at any given point
```{r}
fit_and_predict <- function(.x, ks){
  set.seed(8)
  m1 <- fitHMM(data=.x, nbStates=4, dist=dist, Par0 = Par0_m1,
               estAngleMean = list(angle=TRUE),
               fixPar=fixPar,
               knownStates = ks,
               #formula = ~julian_day, #state 2 to 3. page 47 of vignette shows how to create the beta matrix. Formula for transition probabilities
               stateNames = stateNames)#nlmPar=list(print.level=2)
  
  .x$step_state <- viterbi(m1)
  
  #step state shows the bird's behavior between points, point state shows the birds behavior at points
  .x$point_state <- map(1:nrow(.x),  function(i){
    if(i == 1){
      return(1) #for the first row, return 1 
      return(1)
    } else if(.x[i,"step_state"] == .x[i-1,"step_state"]){ #is the current step state the same as the last step state? Return the current step state
      return(.x[i,"step_state"])
    } else if(.x[i,"step_state"] == 2 & .x[i-1,"step_state"] == 1){ #is the current step state mig and the previous step pre-mig? Return pre-mig
      return(1) 
    } else if(.x[i,"step_state"] == 2 & .x[i-1,"step_state"] == 3){#is the current step state mig and the previous step stopover? Return stopover
      return(3) 
    } else{ #in all other circumstances, return the current step state
      return(.x[i,"step_state"])
    }
    
  }) %>% unlist()
  
  #step state shows the bird's behavior between points, point state shows the birds behavior at points
  
  #Fix the RI problem birds
  # .x <- .x %>% 
  #   group_by(ID) %>%
  #   mutate(ranks = row_number()) %>%
  #   mutate(point_state = ifelse(ranks == "1" & ID %in% RI_problems, 1, point_state)) %>% 
  #   ungroup() %>% 
  #   dplyr::select(-ranks)
  
  return(list(model = m1, results = .x))
}
```


# and now for the accuracy assessment

```{r}
iterate_ids <- function(id_iter, jul_day_cutoff, rn){
  amwo_hmm_clipped <- amwo_hmm %>% #remove locations for the ID in question after the cutoff date
    filter(!(ID == id_iter & julian_day > jul_day_cutoff))
  
  # Make sure there are enough points to delineate a track
  result <- tryCatch({
    ## Defining starting known states
    #Define those birds for which I should estimate starting states (birds tagged in VA or NJ in this season)
    #Defining "tagged this season" as "tagged less than 2 months (~60 days) before the first location of this seasonal delineation"
    
    starting_state_key <- amwo_hmm_clipped %>% #How many days before the first location of the season was this bird captured?
      left_join(capture_dates) %>% 
      group_by(ID) %>% 
      nest() %>%
      mutate(days_from_capture_to_first_loc = map(.x = data, .f = function(iter_df){
        first_time <- iter_df %>% 
          head(n = 1) %>% 
          pull(time)
        
        cap_date <- iter_df %>% 
          head(n = 1) %>% 
          pull(capture_date)
        
        as.numeric(first_time - cap_date)/24 %>% #diff time is in hours, convert to days
          return()
      }))
    
    starting_state_key <- starting_state_key %>% 
      mutate(estimate_starting_state = if_else((str_detect(ID, pattern = "VA") | str_detect(ID, pattern = "NJ")) & days_from_capture_to_first_loc < 60, true = TRUE, false = FALSE)) %>% 
      dplyr::select(ID, estimate_starting_state) %>% 
      ungroup()
    
    amwo_hmm_clipped <- amwo_hmm_clipped %>% 
      left_join(starting_state_key)
    
    #Assign a starting state for each individual
    #This looks clunky, but it's necessary to maintain a momentuHMMData class
    #Can probably get rid of some exceptions after running the revised known states
    ## rank observations (requires correct sorting), ifelse to covert first observations to state 1 and all others NA
    
    known_state_key <- amwo_hmm_clipped %>%
      group_by(ID) %>%
      mutate(ranks = row_number()) %>% 
      mutate(known = ifelse(estimate_starting_state == FALSE & ranks == 1, yes = 1, no = NA)) %>% #if this is the first row of an ID for which I don't want to estimate starting state, the initial step is 1
      mutate(r1p1s2 = if_else(known == 1 & step > 30.2 & animal_name != "RI-2021-59", true = TRUE, false = FALSE)) %>% #if known == 1 and step >30.2, mark as a "rank 1 point 1 step 2" event to be reclassified in post-processing
      mutate(r1p1s2 = replace_na(r1p1s2, replace = FALSE)) %>% #deals with NA issues 
      mutate(known = ifelse(step > 30.2, 2, known)) %>% #if step length is >30.2, then it must be mig (even if known is already 1) 
      mutate(known = ifelse(ID == "RI-2020-42-2020" & julian_day == 331, 2, known)) %>% #fix a bird which goes into 4 too early
      mutate(known = ifelse(ID == "VA-2019-48-2019" & julian_day == 300, 4, known)) %>% #fix a bird which settles after 2/25
      mutate(known = ifelse(ID == "RI-2021-59-2021" & ranks == 1, 3, known)) %>% 
      # mutate(known = ifelse(known == 1 & ID %in% RI_problems, 2, known)) %>%
      group_modify(.f = function(x, y){ #if we know that the bird has a spring migration after this, it must settle
        max_rank <- x %>% 
          pull(ranks) %>% 
          max()
        if(pull(y, ID) %in% spring_init_dates$ID){
          x <- x %>% 
            mutate(known = ifelse(ranks == max_rank, 4, known)) #set the state of the last row to settled
        }
        return(x)
      }) %>% 
      ungroup()
    
    amwo_hmm_clipped$known <- known_state_key$known
    amwo_hmm_clipped$r1p1s2 <- known_state_key$r1p1s2
    
    knownStates <- amwo_hmm_clipped$known
    
    # Running the model
    ##
    results_modified <- fit_and_predict(amwo_hmm_clipped, ks = knownStates)
    ##
    
    #retrieve the modified ending state
    end_state_modified <- results_modified$results %>% 
      filter(ID == id_iter & julian_day == jul_day_cutoff) %>% 
      pull(point_state)
    
    #retrieve the real ending state
    end_state_true <- results %>% 
      filter(ID == id_iter & julian_day == jul_day_cutoff) %>% 
      pull(point_state)
    
    #report the cutoff date, ID, modified ending state, and real ending state in a tibble, which will be the output of the loop
    tibble(cutoff_date = jul_day_cutoff,
           ID = id_iter,
           end_modified = end_state_modified,
           end_true = end_state_true)
  },
  error = function(e){
    tibble(cutoff_date = jul_day_cutoff,
           ID = id_iter,
           end_modified = NA,
           end_true = NA)
  })
  return(result)
}
```

```{r}
cutoffs <- c("10/1/2020", "10/15/2020", "11/1/2020", "11/15/2020", "12/1/2020", "12/15/2020", "1/1/2021", "1/15/2021", "2/1/2021", "2/15/2021") %>% 
  mdy(.) - days_subtract

cutoffs <- yday(cutoffs)

#now iterating through individuals
set.seed(8)
id_subsample <- id_list %>% sample(size = 52) #note: this is a reduced size

parameters_to_run <- expand.grid(id_subsample, cutoffs) %>%
  mutate(rn = row_number())
colnames(parameters_to_run) <- c("id_iter", "jul_day_cutoff", "rn")

parameters_to_run <- parameters_to_run %>% #if I omit the bird from the model, omit it from the accuracy evaluation
  filter(!(id_iter %in% c("PA-2019-13-2019", "RI-2019-18-2019", "RI-2019-29-2019", "VT-2020-04-2020", "FL-2021-01-2021", "NJ-2018-11-2018", "SC-2020-13-2020", "GA-2021-18-2021"))) %>% 
  filter(!(id_iter %in% c("RI-2020-42-2020", "VA-2019-48-2019", "RI-2021-58-2021", "NY-2019-11-2019"))) #also omit birds with manually-set known termination states
```

Take out iterations that would result in fewer than 1 rows for the bird of concern, birds that don't have a location on the cutoff date, and eliminate iterations that I've already run
```{r}
parameters_to_run$valid <- parameters_to_run %>% 
  dplyr::select(id_iter, jul_day_cutoff) %>% 
  pmap(.f = function(id_iter, jul_day_cutoff){
    amwo_hmm_clipped <- amwo_hmm %>% #remove locations for the ID in question after the cutoff date
      filter(ID == id_iter & julian_day < jul_day_cutoff)
    
    observed_on_cutff <- amwo_hmm %>% #ensure that there's a sighting on the cutoff date
      filter(ID == id_iter & julian_day == jul_day_cutoff) %>% 
      nrow()
    
    logic_flag_1 <- nrow(amwo_hmm_clipped) > 1
    
    logic_flag_2 <- observed_on_cutff != 0
    
    if_else(logic_flag_1 & logic_flag_2, TRUE, FALSE) %>% 
      return()
  })

parameters_to_run <- parameters_to_run %>% 
  filter(valid == TRUE) %>% 
  dplyr::select(-valid) #%>% 
  #filter(!(id_iter == "QUE-2019-08-2019" & jul_day_cutoff == 219))

prior_results <- read.csv("accuracy_results_fall_null.csv") %>% #ensure that I'm not duplicating past labor
  transmute(combined = paste0(cutoff_date, "_",ID))

#How many more samples do I need?
read.csv("accuracy_results_fall_null.csv") %>% 
  pull(ID) %>% 
  unique() %>% 
  length()

parameters_to_run <- parameters_to_run %>%
  mutate(combined = paste0(jul_day_cutoff, "_",id_iter)) %>%
  filter(!(combined %in% prior_results$combined)) %>%
  mutate(combined = NULL)

#How many more samples will this give me?
parameters_to_run %>% 
  pull(id_iter) %>% 
  unique() %>% 
  length()
```

Run the process in parallel
```{r}
plan(multisession, workers = 6)
tbl_list <- future_pmap(.l = parameters_to_run, 
                        .f = iterate_ids, 
                        .options = furrr_options(seed = 8),
                        .progress = TRUE)
plan(sequential)
```

```{r}
accuracy_results_fall <- tbl_list %>% 
  do.call(rbind, .)

# Append results to previous runs
write.table(accuracy_results_fall, "accuracy_results_fall_null.csv",
            append = TRUE,
            sep = ",",
            col.names = FALSE,
            row.names = FALSE,
            quote = FALSE)
```

# Read the results back in and plot
```{r}
accuracy_results_fall <- read_csv("accuracy_results_fall_null.csv") %>% 
  filter(!is.na(end_modified))
```

Coding the results in a ggplot-friendly format
```{r}
accuracy_results_fall <- accuracy_results_fall %>% 
  mutate(false_mig = if_else((end_modified == 2 | end_modified == 3) & (end_true == 4 | end_true == 5), 1, 0)) %>% 
  mutate(false_postmig = if_else((end_true == 2 | end_true == 3) & (end_modified == 4 | end_modified == 5), 1, 0)) %>% 
  pivot_longer(cols = c("false_mig", "false_postmig"), names_to = "test", values_to = "result") %>% 
  group_by(cutoff_date, test) %>% 
  summarise(percent_error = mean(result), std_err = plotrix::std.error(result)) %>% 
  ungroup()
```

```{r}
ggplot(data = accuracy_results_fall, mapping = aes(x = cutoff_date, y = percent_error, col = test)) +
  geom_point(position=position_dodge(width=4)) +
  geom_errorbar(aes(ymin = percent_error - std_err*1.96, ymax = percent_error + std_err*1.96), width = 2,
                position=position_dodge(width=4),) +
  theme_bw() +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Percent error", x = "Julian day", col = "Classification error")

ggsave("fall_null_classification_error_DRAFT_5_22_23.png", width = 7, height = 5)
```

