---
title: "R Notebook"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---
```{r include=FALSE}
library(tidyverse)
library(lubridate)
library(move)
library(momentuHMM)
library(leaflet)
library(RColorBrewer)
library(sf)
library(furrr)
library(here)
library(crawl)
```

Load data for use in the assessment
```{r}
load(file=here("classifier_fall", "fall_all_data.RData"))
```

```{r}
# RI_problems <- c("RI-2020-33-2020", "RI-2020-34-2020", "RI-2020-35-2020", "RI-2020-42-2020", "RI-2020-43-2020", "RI-2020-44-2020", "RI-2021-48-2021", "RI-2021-49-2021", "RI-2020-30-2020", "RI-2020-32-2020", "SC-2020-13-2020", "VT-2020-14-2020")
```

Determine which individuals *didn't* die too early (here defined as 12/15, probably need to bump this up to 1/15)
```{r}
id_list <- unique(amwo_hmm$ID)
id_list <- id_list[!(id_list %in% individuals_died_too_early)]
```

Load the results object
```{r}
results <- readRDS(here("classifier_fall", "fall_all_4_state_model.rds"))
```

Frontmatter: prepare the model

Label states and create distributions for movements
```{r}
stateNames <- c("pre","migration", "stopover", "post")
dist <- list(step="gamma", angle="wrpcauchy", y = "norm", julian_day = "norm", dist_start = "bern", step_500 = "bern", breeding_abundance = "bern", log_mean_dist_7 = "norm", residence_time = "norm")
```

fix step length parameters and state transitions
```{r}
fixPar <- list(beta=matrix(c(NA, -1000, -1000, # pre-migration to transitions, can't go pre- to post
                             -1000, NA, NA,  #once entered migration can't go back to pre
                             -1000, NA, -1000, #Once entered stopover, can only go back to migration
                             -1000, -1000, -1000), nrow = 1, byrow = TRUE), #once entered post-migration, can't leave
               step_500 = c(0.000000001, 0.999999999, 0.000000001, 0.000000001)) 
```

These serve as the starting parameters for the HMM to optimize.
```{r}
Par0_m1 <- list(step=c(0.261500539, 3.838994e+02, 0.834030041, 0.261825163, #mean in km
                       0.404219742, 4.410903e+02, 1.488383758, 0.425347015, # SD in km
                       0.003181082, 3.196105e-67, 0.003409205, 0.001801455), #zeromass ie how many zeroes expected in the distribution
                angle=c(-0.000203342, -0.07766199, 0.04102217, -0.001115973, #mean
                        0.9361437423, 0.27836443, 0.25968012, 0.827444744), #concentration
                y = c(42.862454, 39.823877, 37.643459, 33.899562,
                      3.101288, 4.254915, 3.484837, 2.230976),
                julian_day = c(231.3170, 266.31026, 279.18957, 316.30196, 
                               25.4405, 24.72308, 24.87789, 24.78954),
                log_mean_dist_7 = c(4.062038, 7.953790, 4.951820, 3.563583, #mean
                                    1.593476, 3.889766, 2.663651, 1.548703),
                residence_time = c(56.03299, 19.56778, 30.27572, 72.30254, #mean
                                   28.02860, 24.80717, 20.52254, 19.56757),
                dist_start = c(0.003826158, 0.6588471, 0.9374171, 0.9968719), #prob
                step_500 = c(0.00001, 0.99999, 0.00001, 0.00001), #prob
                breeding_abundance = c(0.003084936, 0.2362899, 0.4301947, 0.6869913)) #sd
# lat_diff_abs = c(0.01, 1, 0.01,
#                  0.01, 1, 0.01,
#                  0.026, 0.036, 0.026)) 
```

Create a function that fits the model, predicts states for each step (predicted and observed) and input in the dataset. Note: the listed x y is the starting location of the step. Then use this to determine what the bird's state was at any given point
```{r}
fit_and_predict <- function(.x, ks){
  set.seed(8)
  m1 <- fitHMM(data=.x, nbStates=4, dist=dist, Par0 = Par0_m1,
               estAngleMean = list(angle=TRUE),
               fixPar=fixPar,
               knownStates = ks,
               #formula = ~julian_day, #state 2 to 3. page 47 of vignette shows how to create the beta matrix. Formula for transition probabilities
               stateNames = stateNames)#nlmPar=list(print.level=2)
  
  .x$step_state <- viterbi(m1)
  
  #step state shows the bird's behavior between points, point state shows the birds behavior at points
  .x$point_state <- map(1:nrow(.x),  function(i){
    if(i == 1){
      return(1) #for the first row, return 1 
      return(1)
    } else if(.x[i,"step_state"] == .x[i-1,"step_state"]){ #is the current step state the same as the last step state? Return the current step state
      return(.x[i,"step_state"])
    } else if(.x[i,"step_state"] == 2 & .x[i-1,"step_state"] == 1){ #is the current step state mig and the previous step pre-mig? Return pre-mig
      return(1) 
    } else if(.x[i,"step_state"] == 2 & .x[i-1,"step_state"] == 3){#is the current step state mig and the previous step stopover? Return stopover
      return(3) 
    } else{ #in all other circumstances, return the current step state
      return(.x[i,"step_state"])
    }
    
  }) %>% unlist()
  
  #step state shows the bird's behavior between points, point state shows the birds behavior at points
  
  #Fix the RI problem birds
  # .x <- .x %>% 
  #   group_by(ID) %>%
  #   mutate(ranks = row_number()) %>%
  #   mutate(point_state = ifelse(ranks == "1" & ID %in% RI_problems, 1, point_state)) %>% 
  #   ungroup() %>% 
  #   dplyr::select(-ranks)
  
  return(list(model = m1, results = .x))
}
```


# and now for the accuracy assessment

```{r}
iterate_ids <- function(id_iter, jul_day_cutoff, rn){
  amwo_hmm_clipped <- amwo_hmm %>% #remove locations for the ID in question after the cutoff date
    filter(!(ID == id_iter & julian_day > jul_day_cutoff))
  
  ## Defining starting known states
  #Define those birds for which I should estimate starting states (birds tagged in VA or NJ in this season)
  #Defining "tagged this season" as "tagged less than 2 months (~60 days) before the first location of this seasonal delineation"
  
  starting_state_key <- amwo_hmm_clipped %>% #How many days before the first location of the season was this bird captured?
    left_join(capture_dates) %>% 
    group_by(ID) %>% 
    nest() %>%
    mutate(days_from_capture_to_first_loc = map(.x = data, .f = function(iter_df){
      first_time <- iter_df %>% 
        head(n = 1) %>% 
        pull(time)
      
      cap_date <- iter_df %>% 
        head(n = 1) %>% 
        pull(capture_date)
      
      as.numeric(first_time - cap_date)/24 %>% #diff time is in hours, convert to days
        return()
    }))
  
  starting_state_key <- starting_state_key %>% 
    mutate(estimate_starting_state = if_else((str_detect(ID, pattern = "VA") | str_detect(ID, pattern = "NJ")) & days_from_capture_to_first_loc < 60, true = TRUE, false = FALSE)) %>% 
    dplyr::select(ID, estimate_starting_state) %>% 
    ungroup()
  
  amwo_hmm_clipped <- amwo_hmm_clipped %>% 
    left_join(starting_state_key)
  
  #Assign a starting state for each individual
  #This looks clunky, but it's necessary to maintain a momentuHMMData class
  #Can probably get rid of some exceptions after running the revised known states
  ## rank observations (requires correct sorting), ifelse to covert first observations to state 1 and all others NA
  
  known_state_key <- amwo_hmm_clipped %>%
    group_by(ID) %>%
    mutate(ranks = row_number()) %>% 
    mutate(known = ifelse(estimate_starting_state == FALSE & ranks == 1, yes = 1, no = NA)) %>% #if this is the first row of an ID for which I don't want to estimate starting state, the initial step is 1
    mutate(r1p1s2 = if_else(known == 1 & step > 30.2 & animal_name != "RI-2021-59", true = TRUE, false = FALSE)) %>% #if known == 1 and step >30.2, mark as a "rank 1 point 1 step 2" event to be reclassified in post-processing
    mutate(r1p1s2 = replace_na(r1p1s2, replace = FALSE)) %>% #deals with NA issues 
    mutate(known = ifelse(step > 30.2, 2, known)) %>% #if step length is >30.2, then it must be mig (even if known is already 1) 
    mutate(known = ifelse(ID == "RI-2020-42-2020" & julian_day == 331, 2, known)) %>% #fix a bird which goes into 4 too early
    mutate(known = ifelse(ID == "VA-2019-48-2019" & julian_day == 300, 4, known)) %>% #fix a bird which settles after 2/25
    mutate(known = ifelse(ID == "RI-2021-59-2021" & ranks == 1, 3, known)) %>% 
    # mutate(known = ifelse(known == 1 & ID %in% RI_problems, 2, known)) %>%
    group_modify(.f = function(x, y){ #if we know that the bird has a spring migration after this, it must settle
      max_rank <- x %>% 
        pull(ranks) %>% 
        max()
      if(pull(y, ID) %in% spring_init_dates$ID){
        x <- x %>% 
          mutate(known = ifelse(ranks == max_rank, 4, known)) #set the state of the last row to settled
      }
      return(x)
    }) %>% 
    ungroup()
  
  amwo_hmm_clipped$known <- known_state_key$known
  amwo_hmm_clipped$r1p1s2 <- known_state_key$r1p1s2
  
  knownStates <- amwo_hmm_clipped$known
  
  # Running the model
  ##
  results_modified <- fit_and_predict(amwo_hmm_clipped, ks = knownStates)
  ##
  
  #retrieve the modified ending state
  end_state_modified <- results_modified$results %>% 
    filter(ID == id_iter & julian_day == jul_day_cutoff) %>% 
    pull(point_state)
  
  #retrieve the real ending state
  end_state_true <- results %>% 
    filter(ID == id_iter & julian_day == jul_day_cutoff) %>% 
    pull(point_state)
  
  #report the cutoff date, ID, modified ending state, and real ending state in a tibble, which will be the output of the loop
  tibble(cutoff_date = jul_day_cutoff,
         ID = id_iter,
         end_modified = end_state_modified,
         end_true = end_state_true) %>% 
    return()
}
```

```{r}
cutoffs <- c("10/1/2020", "10/15/2020", "11/1/2020", "11/15/2020", "12/1/2020", "12/15/2020", "1/1/2021", "1/15/2021", "2/1/2021", "2/15/2021") %>% 
  mdy(.) - days_subtract

cutoffs <- yday(cutoffs)

#now iterating through individuals
set.seed(8)
id_subsample <- id_list %>% sample(size = 50)

parameters_to_run <- expand.grid(id_subsample, cutoffs) %>%
  mutate(rn = row_number())
colnames(parameters_to_run) <- c("id_iter", "jul_day_cutoff", "rn")

parameters_to_run <- parameters_to_run %>% #if I omit the bird from the model, omit it from the accuracy evaluation
  filter(!(id_iter %in% c("PA-2019-13-2019", "RI-2019-18-2019", "RI-2019-29-2019", "VT-2020-04-2020", "FL-2021-01-2021", "NJ-2018-11-2018", "SC-2020-13-2020", "GA-2021-18-2021")))
```

Take out iterations that would result in fewer than 1 rows for the bird of concern, and eliminate iterations that I've already run
```{r}
parameters_to_run$valid <- parameters_to_run %>% 
  dplyr::select(id_iter, jul_day_cutoff) %>% 
  pmap(.f = function(id_iter, jul_day_cutoff){
    amwo_hmm_clipped <- amwo_hmm %>% #remove locations for the ID in question after the cutoff date
      filter(ID == id_iter & julian_day < jul_day_cutoff)
    
    if_else(nrow(amwo_hmm_clipped) > 1, TRUE, FALSE) %>% 
      return()
  })

parameters_to_run <- parameters_to_run %>% 
  filter(valid == TRUE) %>% 
  dplyr::select(-valid)

# prior_results <- read.csv("accuracy_results_fall.csv") %>% #ensure that I'm not duplicating past labor
#   transmute(combined = paste0(cutoff_date, "_",ID))
# 
# parameters_to_run <- parameters_to_run %>% 
#   mutate(combined = paste0(jul_day_cutoff, "_",id_iter)) %>% 
#   filter(!(combined %in% prior_results$combined)) %>% 
#   mutate(combined = NULL)
```

Run the process in parallel
```{r}
plan(multisession)
tbl_list <- future_pmap(.l = parameters_to_run, 
                        .f = iterate_ids, 
                        .options = furrr_options(seed = 8),
                        .progress = TRUE)
plan(sequential)
```

```{r}
accuracy_results_fall <- tbl_list %>% 
  do.call(rbind, .)

# Append results to previous runs
write.table(accuracy_results_fall, "accuracy_results_fall_full.csv",
            append = TRUE,
            sep = ",",
            col.names = FALSE,
            row.names = FALSE,
            quote = FALSE)
```

# Read the results back in and plot
```{r}
accuracy_results_fall <- read_csv("accuracy_results_fall_full.csv")
```

Coding the results in a ggplot-friendly format
```{r}
accuracy_results_fall <- accuracy_results_fall %>% 
  mutate(false_mig = if_else((end_modified == 2 | end_modified == 3) & (end_true == 4 | end_true == 5), 1, 0)) %>% 
  mutate(false_postmig = if_else((end_true == 2 | end_true == 3) & (end_modified == 4 | end_modified == 5), 1, 0)) %>% 
  pivot_longer(cols = c("false_mig", "false_postmig"), names_to = "test", values_to = "result") %>% 
  group_by(cutoff_date, test) %>% 
  summarise(percent_error = mean(result), std_err = plotrix::std.error(result)) %>% 
  ungroup()
```

```{r}
ggplot(data = accuracy_results_fall, mapping = aes(x = cutoff_date, y = percent_error, col = test)) +
  geom_point(position=position_dodge(width=4)) +
  geom_errorbar(aes(ymin = percent_error - std_err*1.96, ymax = percent_error + std_err*1.96), width = 2,
                position=position_dodge(width=4),) +
  theme_bw() +
  scale_y_continuous(labels = scales::percent) +
  labs(y = "Percent error", x = "Julian day", col = "Classification error")

ggsave("fall_classification_error_2_22_23.png", width = 7, height = 5)
```

