---
title: "R Notebook"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---
```{r include=FALSE}
library(tidyverse)
library(lubridate)
library(move)
library(momentuHMM)
library(leaflet)
library(RColorBrewer)
library(sf)
library(furrr)
library(here)
library(crawl)
```

Load data for use in the assessment
```{r}
load(file=here("classifier_spring", "spring_female_data.RData"))
```

Determine which individuals *didn't* die too early (here defined as 5/15)
```{r}
id_list <- unique(amwo_hmm$ID)
id_list <- id_list[!(id_list %in% individuals_died_too_early)]
```

Load the results object
```{r}
results <- readRDS(here("classifier_spring", "spring_female_4_state_model.rds"))
```

Frontmatter: prepare the model

Label states and create distributions for movements
```{r}
stateNames <- c("pre", "migration", "stopover", "post")
dist <- list(step="gamma", angle="wrpcauchy", y = "norm", julian_day = "norm", log_mean_dist_7 = "norm", residence_time = "norm", dist_start = "bern", step_500 = "bern", breeding_abundance = "bern") #, lat_diff_abs = "gamma" , pts_10mi_log = "norm"
```

fix step length parameters and state transitions
```{r}
fixPar <- list(beta=matrix(c(NA, -1000,	-1000,	-1000,	NA,	NA,	-1000,	NA,	-1000,	-1000,	-1000,	-1000), nrow = 1, byrow = TRUE),
               step_500 = c(0.000000001, 0.999999999, 0.000000001, 0.000000001))
```

These serve as the starting parameters for the HMM to optimize.
```{r}
Par0_m1 <- list(step=c(0.215581430, 2.327105e+02, 0.765139476, 0.165041155, #mean in km
                       0.346142137, 2.101144e+02, 1.510767790, 0.279555846, # SD in km
                       0.002934171, 1.566103e-09, 0.006099425, 0.009866825), #zeromass ie how many zeroes expected in the distribution
                angle=c(0.001519699, -0.09475137, -0.006933088, 0.0007521754, #mean
                        0.852843337, 0.33560275, 0.379823754, 0.7379986808),
                step_500 = c(1e-09, 0.99999, 1e-09, 1e-09)) #prob
```

Create a function that fits the model, predicts states for each step (predicted and observed) and input in the dataset. Note: the listed x y is the starting location of the step. Then use this to determine what the bird's state was at any given point
```{r}
fit_and_predict <- function(.x, ks, vtt, ms){
  dist_iter <- dist[ms]
  Par0_m1_iter <- Par0_m1[ms]
  
  set.seed(8)
  m1 <- fitHMM(data=.x, nbStates=4, dist=dist_iter, Par0 = Par0_m1_iter,
               estAngleMean = list(angle=TRUE),
               fixPar=fixPar,
               #userBounds = userBounds,
               #workBounds = workBounds,
               #DM = DM,
               knownStates = ks,
               #formula = ~step_cov, #state 2 to 3. page 47 of vignette shows how to create the beta matrix
               stateNames = stateNames)
  
  .x$step_state <- viterbi(m1)
  
  #step state shows the bird's behavior between points, point state shows the birds behavior at points
  .x$point_state <- map(1:nrow(.x),  function(i){
    if(i == 1){
      return(1) #for the first row, return 1
    } else if(.x[i,"ID"] != .x[i-1,"ID"]){ #is this the first point for this ID? Return the starting state
      return(1)
    } else if(.x[i,"step_state"] == .x[i-1,"step_state"]){ #is the current step state the same as the last step state? Return the current step state
      return(.x[i,"step_state"])
    } else if(.x[i,"step_state"] == 2 & .x[i-1,"step_state"] == 1){ #is the current step state mig and the previous step pre-mig? Return pre-mig
      return(1)
    } else if(.x[i,"step_state"] == 2 & .x[i-1,"step_state"] == 3){#is the current step state mig and the previous step stopover? Return stopover
      return(3)
    } else{ #in all other circumstances, return the current step state
      return(.x[i,"step_state"])
    }
    
  }) %>% unlist()
  
  return(list(model = m1, results = .x))
}
```


# and now for the accuracy assessment

```{r}
iterate_ids_male <- function(jul_day_cutoff, variable_to_test, model_structure, fold_name, fold_indicies, to_constrain, data_constrained, true_endings){#id_iter, jul_day_cutoff, rn, variable_to_test, model_structure
  amwo_hmm_clipped <- data_constrained
  
  result <- tryCatch({
    # Assigning starting state
    starting_state_key <- amwo_hmm_clipped %>% #How many days before the first location of the season was this bird captured?
      left_join(capture_dates) %>% 
      group_by(ID) %>% 
      nest() %>%
      mutate(days_from_capture_to_first_loc = map(.x = data, .f = function(iter_df){
        first_time <- iter_df %>% 
          head(n = 1) %>% 
          pull(time)
        
        cap_date <- iter_df %>% 
          head(n = 1) %>% 
          pull(capture_date)
        
        as.numeric(first_time - cap_date)/24 %>% #diff time is in hours, convert to days
          return()
      }))
    
    starting_state_key <- starting_state_key %>% 
      mutate(estimate_starting_state = if_else((str_detect(ID, pattern = "VA") | str_detect(ID, pattern = "NJ") | str_detect(ID, pattern = "MD")) & days_from_capture_to_first_loc < 60, true = TRUE, false = FALSE)) %>% 
      dplyr::select(ID, estimate_starting_state) %>% 
      ungroup()
    
    amwo_hmm_clipped <- amwo_hmm_clipped %>% 
      left_join(starting_state_key)
    
    known_state_key <- amwo_hmm_clipped %>%
      group_by(ID) %>%
      mutate(ranks = row_number()) %>% # rank observations (requires correct sorting), ifelse to convert first observations to state 1 and all others NA
      ungroup() %>% 
      mutate(known = ifelse(estimate_starting_state == FALSE & ranks == 1, yes = 1, no = NA)) %>%
      mutate(r1p1s2 = if_else(known == 1 & step >= 16.1, true = TRUE, false = FALSE)) %>% #if known == 1 and step >30.2, mark as a "rank 1 point 1 step 2" event to be reclassified in post-processing
      mutate(r1p1s2 = replace_na(r1p1s2, replace = FALSE)) %>% #deals with NA issues
      mutate(known = ifelse(step >= 16.1, 2, known)) %>% #if step length is >30.2, then it must be mig (even if known is already 1) 
      mutate(known = ifelse(ID == "VA-2021-91-2021" & ranks %in% c(16, 17, 18), 2, known)) %>% #crw stops the bird from slipping into mig
      mutate(known = ifelse(ID == "VA-2021-92-2021" & ranks %in% c(6, 7, 8), 2, known)) %>% #crw stops the bird from slipping into mig
      mutate(known = ifelse(ID == "NJ-2018-15-2019" & julian_day == 46, 2, known)) %>% 
      mutate(known = ifelse(ID == "VA-2019-22-2019" & julian_day == 99, 2, known)) %>% 
      mutate(known = ifelse(ID == "ME-2018-13-2019" & julian_day == 161, 2, known)) %>% 
      mutate(known = ifelse(ID == "NY-2022-40-2022" & ranks == 1, 3, known)) %>%  #bird that migrated after nesting starts in stopover state
      mutate(known = ifelse(ID == "RI-2018-11-2019" & ranks == 95, 4, known)) #bird is recaptured and re-transmittered at the last mig location the following Oct, so we know that it is settled here
    
    amwo_hmm_clipped$known <- known_state_key$known
    amwo_hmm_clipped$r1p1s2 <- known_state_key$r1p1s2
    
    knownStates <- amwo_hmm_clipped$known
    
    # Running the model
    results_modified <- fit_and_predict(amwo_hmm_clipped, ks = knownStates, vtt = variable_to_test, ms = model_structure)
    
    #retrieve the modified ending state
    
    end_state_modified <- numeric()
    
    for(i in 1:nrow(to_constrain)){
      id_to_clip <- to_constrain$id_iter[i]
      dummy_y <- jul_day_cutoff
      
      j <- results_modified$results %>% 
        filter(ID == id_to_clip & julian_day == dummy_y) %>% 
        head(n=1) %>% 
        pull(point_state)
      
      end_state_modified <- append(end_state_modified, j)
    }
    
    if(length(end_state_modified)!=nrow(to_constrain)){
      stop("Mismatch between number of IDs and length of output")
    }
    
    # end_state_modified <- results_modified$results %>% 
    #   filter(ID == id_iter & julian_day == jul_day_cutoff) %>% 
    #   pull(point_state)
    
    #report the cutoff date, ID, modified ending state, and real ending state in a tibble, which will be the output of the loop
    tibble(jul_day_cutoff = jul_day_cutoff,
           variable_to_test = variable_to_test,
           model_structure = list(model_structure),
           fold_name = fold_name,
           fold_indicies = list(fold_indicies),
           to_constrain = list(to_constrain),
           data_constrained = list(data_constrained),
           end_modified = list(end_state_modified),
           end_true = list(true_endings))
  },
  error = function(e){
    tibble(jul_day_cutoff = jul_day_cutoff,
           variable_to_test = variable_to_test,
           model_structure = list(model_structure),
           fold_name = fold_name,
           fold_indicies = list(fold_indicies),
           to_constrain = list(to_constrain),
           data_constrained = list(data_constrained),
           end_modified = list(NA),
           end_true = list(true_endings))
  })
  return(result)
}
```


```{r}
cutoffs <- c("3/15/2020", "4/1/2020", "4/15/2020", "5/1/2020", "5/15/2020", "6/1/2020", "6/15/2020") %>% 
  mdy() %>% 
  yday()

#now iterating through individuals
set.seed(8)
id_subsample <- id_list %>% sample(size = 50)

parameters_to_run <- expand.grid(id_subsample, cutoffs) %>%
  mutate(rn = row_number())
colnames(parameters_to_run) <- c("id_iter", "jul_day_cutoff", "rn")

parameters_to_run <- parameters_to_run %>% #if I omit the bird from the model, omit it from the accuracy evaluation
  filter(!(id_iter %in% c("NJ-2019-28-2020", "NS-2019-02-2020", "RI-2020-42-2021", "RI-2021-47-2022", "RI-2021-52-2022"))) %>% 
  filter(!(id_iter %in% c("RI-2018-11-2019")))# omit birds that needed to be set as settled manually
```

```{r}
parameters_to_run$valid <- parameters_to_run %>% 
  dplyr::select(id_iter, jul_day_cutoff) %>% 
  pmap(.f = function(id_iter, jul_day_cutoff){
    amwo_hmm_clipped <- amwo_hmm %>% #remove locations for the ID in question after the cutoff date
      filter(ID == id_iter & julian_day < jul_day_cutoff)
    
    observed_on_cutff <- amwo_hmm %>% #ensure that there's a sighting on the cutoff date
      filter(ID == id_iter & julian_day == jul_day_cutoff) %>% 
      nrow()
    
    logic_flag_1 <- nrow(amwo_hmm_clipped) > 1
    
    logic_flag_2 <- observed_on_cutff != 0
    
    if_else(logic_flag_1 & logic_flag_2, TRUE, FALSE) %>% 
      return()
  })

parameters_to_run <- parameters_to_run %>% 
  filter(valid == TRUE) %>% 
  dplyr::select(-valid)

#How many more samples will this give me?
parameters_to_run %>% 
  pull(id_iter) %>% 
  unique() %>% 
  length()

parameters_to_run$model_structure <- map(.x = NA, function(x){
  c("step", "angle", "step_500")
})

parameters_to_run$variable_to_test <- "null"
```

Create a nested version of parameters_to_run that will measure multiple errors at once, in a testing/training fold framework
```{r}
parameters_nested <- parameters_to_run %>% 
  group_by(variable_to_test, model_structure, jul_day_cutoff) %>% 
  nest(.key = "ids")

parameters_nested$folds <- map2(.x = parameters_nested$ids, .y = 1:nrow(parameters_nested), .f = function(x, y){
  nestfs::create.folds(num.folds = 5, num.rows = nrow(x), seed = y)
})

parameters_nested <- unnest_wider(data = parameters_nested, col = folds, names_sep = "") %>% 
  pivot_longer(cols = folds1:folds5, names_to = "fold_name", values_to = "fold_indicies")

#the constraints used for each iteration of the model
parameters_nested$to_constrain <- map2(.x = parameters_nested$ids, .y = parameters_nested$fold_indicies, .f = function(x, y){
  x[y,] %>% 
    mutate(rn = NULL) %>% 
    return()
})

parameters_nested$ids <- NULL #this has been replaced by to_constrain

#amwo locations filtered to the constraints set
parameters_nested$data_constrained <- map2(.x = parameters_nested$to_constrain, .y = parameters_nested$jul_day_cutoff, .f = function(x, y){
  
  amwo_hmm_clipped <- amwo_hmm
  
  for(i in 1:nrow(x)){
    id_to_clip <- x$id_iter[i]
    dummy_y <- y
    
    amwo_hmm_clipped <- amwo_hmm_clipped %>% #remove locations for the ID in question after the cutoff date
      filter(!(ID == id_to_clip & julian_day > dummy_y))
  }
  return(amwo_hmm_clipped)
})

parameters_nested$true_endings <- map2(.x = parameters_nested$to_constrain, .y = parameters_nested$jul_day_cutoff, .f = function(x, y){
  
  trgt <- numeric()
  
  for(i in 1:nrow(x)){
    id_to_clip <- x$id_iter[i]
    dummy_y <- y
    
    j <- results %>% 
      filter(ID == id_to_clip & julian_day == dummy_y) %>%
      head(n=1) %>% 
      pull(point_state)
    
    trgt <- append(trgt, j)
  }
  
  if(length(trgt)!=nrow(x)){
    stop("Mismatch between number of IDs and length of output")
  }
  
  return(trgt)
})
```

```{r}
plan(multisession, workers = 4)
tbl_list <- parameters_nested %>% 
  future_pmap(.l = ., 
                        .f = iterate_ids_male, 
                        .options = furrr_options(seed = 8),
                        .progress = TRUE)
plan(sequential)
```

```{r}
accuracy_results_male <- tbl_list %>% 
  do.call(rbind, .)

accuracy_results_male %>% 
  dplyr::select(-data_constrained) %>% 
  saveRDS("accuracy_results_female_null.rds")
```

