---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(here)
```

# mapping our classification columns -> Movebank attributes

primary_step_state (Migrating/not migrating) -> migration stage custom

primary_point_state (Migrating/not migrating) -> behavioural classification

secondary_step_state (Within-stage ranging movements) -> comments

secondary_point_state (Within-stage ranging movements) -> habitat

nesting -> receiver ID

nocturnal_diurnal -> transmission protocol

flight_ground -> data decoding software

```{r}
amwo_data_reconstructed <- readRDS(here("classifier_integrated", "fac_primary_state_delineation.rds"))

t <- amwo_data_reconstructed %>%
  transmute(event_id = as.character(event_id), 
            primary_step_state = paste("Step state:", primary_step_state), 
            primary_point_state = paste("Point state:", primary_point_state),
            nesting_attempt = nesting_attempt)

t %>%
  write.csv(file = "fac_primary_state_delineation.csv", row.names = FALSE, quote = FALSE)
```
