---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(here)
```

# mapping our classification columns -> Movebank attributes

primary_step_state (Migrating/not migrating) -> migration stage custom

primary_point_state (Migrating/not migrating) -> behavioural classification

nesting -> study-specific-measurement

mortality-status: Binary variable expressing whether a location was recorded after a remotely detected mortality event.

BAS confidence: Binary variable expressing whether a location has a >50% probability of being recorded during a migratory flight.

BAS transition 1 : Probability of a given location being recorded during a migratory flight.

barometric height: Height of a given location above ground level (processed from height-above-msl).

```{r}
amwo_data_reconstructed <- readRDS(here("classifier_integrated", "fac_primary_state_delineation.rds"))

# let's now exclude all individuals which don't have at least one new (undesignated) point
new_animals_to_revise <- amwo_data_reconstructed %>% ### reminder to put this in the movebank upload code...
  filter(designated_before == FALSE) %>% 
  pull(animal_name) %>% 
  unique()

amwo_data_reconstructed <- amwo_data_reconstructed %>% 
  filter(animal_name %in% new_animals_to_revise)

t <- amwo_data_reconstructed %>%
  transmute(event_id = as.character(event_id), 
            primary_step_state = paste("Step state:", primary_step_state), 
            primary_point_state = paste("Point state:", primary_point_state),
            nesting_attempt = nesting_attempt)

t %>%
  write.csv(file = "fac_primary_state_delineation.csv", row.names = FALSE, quote = FALSE)
```
