---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(lubridate)
library(here)
```

Read in data
```{r}
fall_classified <- readRDS(here::here("classifier_fall", "fall_all_4_state_model_base.rds"))
spring_male_classified <- readRDS(here::here("classifier_spring", "spring_male_5_state_model_base.rds"))
spring_female_classified <- readRDS(here::here("classifier_spring", "spring_female_4_state_model_base.rds"))
```

Pulling the migratory initiation and termination date-times
If the starting state != 1, then the initiation date is the date of the first location
```{r}
fall_initiation_dates <- fall_classified %>%
  group_by(ID) %>%
  group_modify(function(x, ...){
    starting_point_state <- x %>% 
      head(n = 1) %>% 
      pull(point_state)
    
    if(starting_point_state == 1){
      x %>% 
        filter(step_state == 2) %>%
        head(x, n = 1) %>%
        mutate(init_state = "Stationary") %>%
        return()
    } else if(starting_point_state == 2 | starting_point_state == 3){
      x %>% 
        head(x, n = 1) %>%
        mutate(init_state = "Migratory") %>%
        return()
    } else{
      stop("Starting point state is not 1, 2, or 3")
    }
  }) %>%
  dplyr::select(ID, animal_name, year, time, init_state) %>%
  rename(time_initiation = time)

fall_termination_dates <- fall_classified %>%
  filter(step_state == 4) %>%
  group_by(ID) %>%
  group_modify(function(x, ...){head(x, n = 1)}) %>%
  dplyr::select(ID, animal_name, year, time) %>%
  rename(time_termination = time)

spring_male_initiation_dates <- spring_male_classified %>%
  group_by(ID) %>%
  group_modify(function(x, ...){
    starting_point_state <- x %>% 
      head(n = 1) %>% 
      pull(point_state)
    
    if(starting_point_state == 1){
      x %>% 
        filter(step_state == 2) %>%
        head(x, n = 1) %>%
        mutate(init_state = "Stationary") %>% 
        return()
    } else if(starting_point_state == 2 | starting_point_state == 3){
      x %>% 
        head(x, n = 1) %>%
        mutate(init_state = "Migratory") %>%
        return()
    } else{
      stop("Starting point state is not 1, 2, or 3")
    }
  }) %>%
  dplyr::select(ID, animal_name, year, time, init_state) %>%
  rename(time_initiation = time) %>% 
  mutate(season = "Spring (male)")

spring_male_termination_dates <- spring_male_classified %>%
  filter(step_state == 4 | step_state == 5) %>%
  group_by(ID) %>%
  group_modify(function(x, ...){head(x, n = 1)}) %>%
  dplyr::select(ID, animal_name, year, time) %>%
  rename(time_termination = time) %>% 
  mutate(season = "Spring (male)")

spring_female_initiation_dates <- spring_female_classified %>%
  group_by(ID) %>%
  group_modify(function(x, ...){
    starting_point_state <- x %>% 
      head(n = 1) %>% 
      pull(point_state)
    
    if(starting_point_state == 1){
      x %>% 
        filter(step_state == 2) %>%
        head(x, n = 1) %>%
        mutate(init_state = "Stationary") %>%
        return()
    } else if(starting_point_state == 2 | starting_point_state == 3){
      x %>% 
        head(x, n = 1) %>%
        mutate(init_state = "Migratory") %>%
        return()
    } else{
      stop("Starting point state is not 1, 2, or 3")
    }
  }) %>%
  dplyr::select(ID, animal_name, year, time, init_state) %>%
  rename(time_initiation = time)  %>% 
  mutate(season = "Spring (female)")

spring_female_termination_dates <- spring_female_classified %>%
  filter(step_state == 4) %>%
  group_by(ID) %>%
  group_modify(function(x, ...){head(x, n = 1)}) %>%
  dplyr::select(ID, animal_name, year, time) %>%
  rename(time_termination = time) %>% 
  mutate(season = "Spring (female)") %>% 
  mutate(time_termination = if_else(animal_name == "VA-2019-45", lubridate::ymd_hms("2020-07-28 19:01:20"), time_termination)) # Manually set a late settlement (after 6-30)

spring_initiation_dates <- bind_rows(spring_male_initiation_dates, spring_female_initiation_dates)
spring_termination_dates <- bind_rows(spring_male_termination_dates, spring_female_termination_dates)

fall_migration_dates <- full_join(fall_initiation_dates, fall_termination_dates) %>%
  ungroup()

spring_migration_dates <- full_join(spring_initiation_dates, spring_termination_dates) %>%
  ungroup()

fall_migration_dates <- fall_migration_dates %>%
  dplyr::select(animal_name, time_initiation, time_termination, init_state) %>%
  mutate(season = "Fall") %>% 
  mutate(time_initiation = as_date(time_initiation),
         time_termination = as_date(time_termination)) %>% 
  mutate(time_initiation = time_initiation - 56,
         time_termination = time_termination - 56) %>% 
  mutate(initiation_standard = mdy(paste0(month(time_initiation), "/", day(time_initiation), "/2020")),
         termination_standard = mdy(paste0(month(time_termination), "/", day(time_termination), "/2020"))) %>% 
  mutate(initiation_standard = initiation_standard + 56,
         termination_standard = termination_standard + 56,
         time_initiation = time_initiation + 56, # undoing the adjustment to put all initiation dates in the same year
         time_termination = time_termination + 56)

spring_migration_dates <- spring_migration_dates %>%
  dplyr::select(animal_name, time_initiation, time_termination, init_state, season) %>%
  mutate(time_initiation = as_date(time_initiation),
         time_termination = as_date(time_termination)) %>% 
  mutate(initiation_standard = mdy(paste0(month(time_initiation), "/", day(time_initiation), "/2020")),
         termination_standard = mdy(paste0(month(time_termination), "/", day(time_termination), "/2020")))

migration_dates <- bind_rows(fall_migration_dates, spring_migration_dates)
```

Fall migration initiation dates
```{r}
fall_initiation_dates <- migration_dates %>% 
  filter(init_state == "Stationary",
         season == "Fall") %>% 
  pull(initiation_standard)

median(fall_initiation_dates)
range(fall_initiation_dates)

#95th quantile of all departure dates
fall_initiation_dates %>% 
  quantile(c(0.05, 0.95), type = 1)
```

Fall migration termination dates
```{r}
fall_termination_dates <- migration_dates %>% 
  filter(season == "Fall") %>% 
  pull(termination_standard)

median(fall_termination_dates, na.rm = TRUE)
range(fall_termination_dates, na.rm = TRUE)

#95th quantile of termination dates
fall_termination_dates %>% 
  quantile(c(0.05, 0.95), type = 1, na.rm = TRUE)
```

Percent completion of fall migration
```{r}
num_completed_fall <- migration_dates %>% 
  filter(season == "Fall") %>% 
  filter(!is.na(time_termination)) %>% 
  nrow()

num_total_fall <- migration_dates %>% 
  filter(season == "Fall") %>% 
  nrow()

num_completed_fall/num_total_fall
```


Spring migration departure dates
```{r}
# Male
spring_male_initiation_dates <- migration_dates %>% 
  filter(init_state == "Stationary",
         season == "Spring (male)") %>% 
  pull(initiation_standard)

print("Male")
median(spring_male_initiation_dates)
range(spring_male_initiation_dates)

#95th quantile of departure dates
spring_male_initiation_dates %>% 
  quantile(c(0.05, 0.95), type = 1)

# Female
spring_female_initiation_dates <- migration_dates %>% 
  filter(init_state == "Stationary",
         season == "Spring (female)") %>% 
  pull(initiation_standard)

print("Female")
median(spring_female_initiation_dates)
range(spring_female_initiation_dates)

#95th quantile of departure dates
spring_female_initiation_dates %>% 
  quantile(c(0.05, 0.95), type = 1)
```

Spring migration termination dates
```{r}
#Male
spring_male_termination_dates <- migration_dates %>% 
  filter(season == "Spring (male)") %>% 
  pull(termination_standard)

print("Male")
median(spring_male_termination_dates, na.rm = TRUE)
range(spring_male_termination_dates, na.rm = TRUE)

#95th quantile of termination dates
spring_male_termination_dates %>% 
  quantile(c(0.05, 0.95), type = 1, na.rm = TRUE)

#Female
spring_female_termination_dates <- migration_dates %>% 
  filter(season == "Spring (female)") %>% 
  pull(termination_standard)

print("Female")
median(spring_female_termination_dates, na.rm = TRUE)
range(spring_female_termination_dates, na.rm = TRUE)

#95th quantile of termination dates
spring_female_termination_dates %>% 
  quantile(c(0.05, 0.95), type = 1, na.rm = TRUE)
```

Percent completion of spring migration
```{r}
#Male 
num_completed_spring_m <- migration_dates %>% 
  filter(season == "Spring (male)") %>% 
  filter(!is.na(time_termination)) %>% 
  nrow()

num_total_spring_m <- migration_dates %>% 
  filter(season == "Spring (male)") %>% 
  nrow()

num_completed_spring_m/num_total_spring_m

#Female
num_completed_spring_f <- migration_dates %>% 
  filter(season == "Spring (female)") %>% 
  filter(!is.na(time_termination)) %>% 
  nrow()

num_total_spring_f <- migration_dates %>% 
  filter(season == "Spring (female)") %>% 
  nrow()

num_completed_spring_f/num_total_spring_f
```