---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(lubridate)
library(here)
```

```{r}
integrated_results <- readRDS(here("classifier_integrated", "fac_primary_state_delineation.rds"))
```

How many birds skipped their fall migrations?
Question to ask- for every year/season in consideration, did the bird have locations before, after, and during that season?
And if so, did it have at least one migratory step during that season?
```{r}
did_you_skip <- function(season_start_date, season_end_date){
  integrated_results %>% 
    group_by(animal_name) %>% 
    group_map(.f = function(x, y){
      locations_before_start <- x %>% 
        filter(time < season_start_date) %>% 
        nrow()
      
      locations_before_start <- locations_before_start > 0
      
      locations_after_end <- x %>% 
        filter(time > season_end_date) %>% 
        nrow()
      
      locations_after_end <- locations_after_end > 0
      
      locations_during_period <- x %>% 
        filter(time > season_start_date & time < season_end_date) %>% 
        nrow()
      
      locations_during_period <- locations_during_period > 0
      
      #Combined logical flag
      active_during_period <- locations_before_start & locations_after_end & locations_during_period
      
      #If active during period, then return whether there was a mig step. Else, return NA
      if(active_during_period){
        all_step_states <- x %>% 
          filter(time > season_start_date & time < season_end_date) %>%
          pull(primary_step_state)
        
        migrated_flag <- ("Migratory (fall)" %in% all_step_states) | ("Migratory (spring)" %in% all_step_states)
        
        tibble(animal_name = y[1,1], active = active_during_period, migrated = migrated_flag) %>% 
          return()
        
      } else {
        tibble(animal_name = y[1,1], active = active_during_period, migrated = NA) %>% 
          return()
      }
    }) %>%
    bind_rows() %>%
    filter(!is.na(migrated)) %>% 
    return()
}
```

Seasons to delineate:
Fall- 2017, 2018, 2019, 2020, and 2021
Spring- 2018, 2019, 2020, 2021, and 2022
```{r}
fall_nonmigs_2017 <- did_you_skip(season_start_date = ymd("2017-10-15"), season_end_date = ymd("2017-12-05")) %>% 
  mutate(season = "Fall", year = 2017)

fall_nonmigs_2018 <- did_you_skip(season_start_date = ymd("2018-10-15"), season_end_date = ymd("2018-12-05")) %>% 
  mutate(season = "Fall", year = 2018)

fall_nonmigs_2019 <- did_you_skip(season_start_date = ymd("2019-10-15"), season_end_date = ymd("2019-12-05")) %>% 
  mutate(season = "Fall", year = 2019)

fall_nonmigs_2020 <- did_you_skip(season_start_date = ymd("2020-10-15"), season_end_date = ymd("2020-12-05")) %>% 
  mutate(season = "Fall", year = 2020)

fall_nonmigs_2021 <- did_you_skip(season_start_date = ymd("2021-10-15"), season_end_date = ymd("2021-12-05")) %>% 
  mutate(season = "Fall", year = 2021)

spring_nonmigs_2018 <- did_you_skip(season_start_date = ymd("2018-2-15"), season_end_date = ymd("2018-3-29")) %>% 
  mutate(season = "Spring", year = 2018)

spring_nonmigs_2019 <- did_you_skip(season_start_date = ymd("2019-2-15"), season_end_date = ymd("2019-3-29")) %>% 
  mutate(season = "Spring", year = 2019)

spring_nonmigs_2020 <- did_you_skip(season_start_date = ymd("2020-2-15"), season_end_date = ymd("2020-3-29")) %>% 
  mutate(season = "Spring", year = 2020)

spring_nonmigs_2021 <- did_you_skip(season_start_date = ymd("2021-2-15"), season_end_date = ymd("2021-3-29")) %>% 
  mutate(season = "Spring", year = 2021)

spring_nonmigs_2022 <- did_you_skip(season_start_date = ymd("2022-2-15"), season_end_date = ymd("2022-3-29")) %>% 
  mutate(season = "Spring", year = 2022)
```

Combine all fall and spring birds together
```{r}
fall_nonmigs <- bind_rows(fall_nonmigs_2017, fall_nonmigs_2018, fall_nonmigs_2019, fall_nonmigs_2020, fall_nonmigs_2021)

spring_nonmigs <- bind_rows(spring_nonmigs_2018, spring_nonmigs_2019, spring_nonmigs_2020, spring_nonmigs_2021, spring_nonmigs_2022)
```

Individually examine each non-migratory bird, verify that it didn't migrate, and determine where.
```{r}
all_nonmigs <- fall_nonmigs %>% 
  bind_rows(spring_nonmigs) %>% 
  filter(migrated == FALSE)
```

n: no data available on other seasons
a1: data available, and migrated in that other season

New:
Selecting from 35 potentially non-mig birds (most of these just migrated late)
F n PA-2019-18 Fall non-migratory PA
M n RI-2019-29 Fall non-migratory RI (w/ dispersal movement)
M a1 VA-2020-52 Fall non-migratory CT (w/ dispersal movement)
M n VA-2020-61 Fall non-migratory VA
F n RI-2021-55 Fall non-migratory RI
F n RI-2021-56 Fall non-migratory RI

F a1 NY-2019-11 Spring non-migratory GA after migrating from NY
M n FL-2021-01 Spring non-migratory FL (w/ foray loop)
F a1 RI-2020-37 Spring non-migratory MD
F a1 RI-2020-40 Spring non-migratory VA
F n VA-2020-73 Spring non-migratory VA
F a1 QUE-2021-25 Spring non-migratory AL after migrating from QUE


Fall tally: 6 of 176 (3.4%)
Locations:
1 PA
3 RI
1 CT
1 VA

Spring tally: 6 of 193 (3.1%)
Locations:
1 GA
1 FL
1 MD
2 VA
1 AL

5 of 5 migrated in another season


Old:

n RI-2018-02 Fall non-migratory RI
n PA-2019-13 Fall non-migratory PA
n PA-2019-18 Fall non-migratory PA
a1 RI-2019-18 Fall non-migratory RI, migrated the next spring
n RI-2019-29 Fall non-migratory RI (w/ dispersal movement)
n VA-2020-61 Fall non-migratory VA
n RI-2021-55 Fall non-migratory RI
n RI-2021-56 Fall non-migratory RI

n FL-2021-01 Spring non-migratory FL (w/ foray loop)
a1 RI-2020-37 Spring non-migratory MD
a1 RI-2020-40 Spring non-migratory VA
n VA-2020-73 Spring non-migratory VA
a1 QUE-2021-25 Spring non-migratory Alabama after migrating from QUE

#had a slight data gap which excludes it from delineation, so excluding
a1 ME-2020-18 Spring non-migratory North Carolina after migrating from ME 

Fall tally: 8 of 174 (5%)
Locations:
RI 5
PA 2
VA 1

Spring tally: 7 of 194 (4%)
Locations:
VA 2
GA 1
FL 1
MD 1
AL 1
NC 1
6 of 6 migrated in another season
