---
title: "R Notebook"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: inline
---

```{r}
library(tidyverse)
library(lubridate)
library(move)
library(momentuHMM)
library(leaflet)
library(RColorBrewer)
library(sf)
library(furrr)
library(here)
library(crawl)
```

Load data for use in the assessment
```{r}
load(file=here("classifier_spring", "spring_male_data.RData"))
```

Determine which individuals *didn't* die too early (here defined as 5/15)
```{r}
id_list <- unique(amwo_hmm$ID)
id_list <- id_list[!(id_list %in% individuals_died_too_early)]
```

Load the results object
```{r}
results <- readRDS(here("classifier_spring", "spring_male_5_state_model.rds"))
```

Frontmatter: prepare the model

Label states and create distributions for movements
```{r}
stateNames <- c("pre", "migration", "stopover", "settling", "post")
dist <- list(step="gamma", angle="wrpcauchy", y = "norm", julian_day = "norm", log_mean_dist_7 = "norm", residence_time = "norm", dist_start = "bern", step_500 = "bern", breeding_abundance = "bern")
```

fix step length parameters and state transitions
```{r}
fixPar <- list(beta=matrix(c(NA,	-10000,	-10000,	-10000,	
                             -10000,	NA,	NA,	NA,	
                             -10000,	NA,	-10000, -10000, 
                             -10000, -10000, -10000, NA, 
                             -10000,	-10000,	-10000,	NA), nrow = 1, byrow = TRUE),
               step_500 = c(0.000000001, 0.999999999, 0.000000001, 0.000000001, 0.000000001)) 
```

These serve as the starting parameters for the HMM to optimize.
```{r}
Par0_m1 <- list(step=c(0.293641627, 79.029506936, 0.157361645, 0.250739740, 0.086056474, #mean in km
                       0.455493251, 95.302307236, 0.211421668, 0.384581337, 0.128558113, # SD in km
                       0.005159307, 0.006024863, 0.006664064, 0.005636388, 0.006973284), #zeromass ie how many zeroes expected in the distribution
                angle=c(0.001353412, 0.0006979447, 0.01168102, 0.0005445842, 0, #mean
                        0.917280917, 0.9754146876, 0.59501513, 0.7614896594, 0.99999), #concentration
                y = c(34.83030, 40.173268, 40.426752,  44.689995, 43.678101,
                      3.11158, 4.497807, 3.668652, 2.267504, 4.109156),
                julian_day = c(40.14297, 79.36682, 80.54528, 135.69949, 136.7710, #mean
                               19.89502, 27.50774, 23.92289, 25.77254, 30.0146),
                log_mean_dist_7 = c(4.126427, 9.965601, 4.959182, 3.797415, 3.581605, #mean
                                    1.679125, 2.403601, 2.736004, 1.536724, 1.806670), #sd
                residence_time = c(39.91826, 6.848351, 53.64592, 71.41284, 70.66562, #mean
                                   20.41741, 13.174650, 35.49282, 21.65228, 25.53273), #sd
                dist_start = c(0.00530222, 0.9276614, 0.9750638, 0.9955032, 0.9882519), #prob
                step_500 = c(0.1336989, 0.99999, 0.06175472, 0.1351401, 0.04079765), #prob
                breeding_abundance = c(0.1183127, 0.3973247, 0.442492, 0.7108834, 0.5959887)) #prob
```


Create a function that fits the model, predicts states for each step (predicted and observed) and input in the dataset. Note: the listed x y is the starting location of the step. Then use this to determine what the bird's state was at any given point
```{r}
fit_and_predict <- function(.x, ks, vtt, ms){
  dist_iter <- dist[ms]
  Par0_m1_iter <- Par0_m1[ms]
  
  set.seed(8)
  m1 <- fitHMM(data=.x, nbStates=5, dist=dist_iter, Par0 = Par0_m1_iter,
               estAngleMean = list(angle=TRUE),
               fixPar=fixPar,
               #userBounds = userBounds,
               #workBounds = workBounds,
               #DM = DM,
               knownStates = ks,
               #formula = ~step_cov, #state 2 to 3. page 47 of vignette shows how to create the beta matrix
               stateNames = stateNames)
  
  .x$step_state <- viterbi(m1)
  
  #step state shows the bird's behavior between points, point state shows the birds behavior at points
  .x$point_state <- map(1:nrow(.x),  function(i){
    if(i == 1){
      return(1) #for the first row, return 1
    } else if(.x[i,"ID"] != .x[i-1,"ID"]){ #is this the first point for this ID? Return the starting state
      return(1)
    } else if(.x[i,"step_state"] == .x[i-1,"step_state"]){ #is the current step state the same as the last step state? Return the current step state
      return(.x[i,"step_state"])
    } else if(.x[i,"step_state"] == 2 & .x[i-1,"step_state"] == 1){ #is the current step state mig and the previous step pre-mig? Return pre-mig
      return(1)
    } else if(.x[i,"step_state"] == 2 & .x[i-1,"step_state"] == 3){#is the current step state mig and the previous step stopover? Return stopover
      return(3)
    } else{ #in all other circumstances, return the current step state
      return(.x[i,"step_state"])
    }
    
  }) %>% unlist()
  
  return(list(model = m1, results = .x))
}
```


# and now for the accuracy assessment

```{r}
iterate_ids_male <- function(id_iter, jul_day_cutoff, rn, variable_to_test, model_structure){
  amwo_hmm_clipped <- amwo_hmm %>% #remove locations for the ID in question after the cutoff date
    filter(!(ID == id_iter & julian_day > jul_day_cutoff))
  
  result <- tryCatch({
    # Assigning starting state
    starting_state_key <- amwo_hmm_clipped %>% #How many days before the first location of the season was this bird captured?
      left_join(capture_dates) %>% 
      group_by(ID) %>% 
      nest() %>%
      mutate(days_from_capture_to_first_loc = map(.x = data, .f = function(iter_df){
        first_time <- iter_df %>% 
          head(n = 1) %>% 
          pull(time)
        
        cap_date <- iter_df %>% 
          head(n = 1) %>% 
          pull(capture_date)
        
        as.numeric(first_time - cap_date)/24 %>% #diff time is in hours, convert to days
          return()
      }))
    
    starting_state_key <- starting_state_key %>% 
      mutate(estimate_starting_state = if_else((str_detect(ID, pattern = "VA") | str_detect(ID, pattern = "NJ") | str_detect(ID, pattern = "MD")) & days_from_capture_to_first_loc < 60, true = TRUE, false = FALSE)) %>% 
      dplyr::select(ID, estimate_starting_state) %>% 
      ungroup()
    
    amwo_hmm_clipped <- amwo_hmm_clipped %>% 
      left_join(starting_state_key)
    
    known_state_key <- known_state_key <- amwo_hmm_clipped %>% 
      group_by(ID) %>% 
      mutate(ranks = row_number()) %>% # rank observations (requires correct sorting), ifelse to convert first observations to state 1 and all others NA
      ungroup() %>%
      mutate(known = ifelse(estimate_starting_state == FALSE & ranks == 1, yes = 1, no = NA)) %>% #if this is the first row of an ID for which I don't want to estimate starting state, the initial step is 1
      mutate(r1p1s2 = if_else(known == 1 & step >= 16.1, true = TRUE, false = FALSE)) %>% #if known == 1 and step >30.2, mark as a "rank 1 point 1 step 2" event to be reclassified in post-processing
      mutate(r1p1s2 = replace_na(r1p1s2, replace = FALSE)) %>% #deals with NA issues 
      mutate(known = ifelse(step >= 16.1, 2, known)) #if step length is >30.2, then it must be mig (even if known is already 1) 
    
    amwo_hmm_clipped$known <- known_state_key$known
    amwo_hmm_clipped$r1p1s2 <- known_state_key$r1p1s2
    
    knownStates <- amwo_hmm_clipped$known
    
    # Running the model
    results_modified <- fit_and_predict(amwo_hmm_clipped, ks = knownStates, vtt = variable_to_test, ms = model_structure)
    
    #retrieve the modified ending state
    end_state_modified <- results_modified$results %>% 
      filter(ID == id_iter & julian_day == jul_day_cutoff) %>% 
      pull(point_state)
    
    #retrieve the real ending state
    end_state_true <- results %>% 
      filter(ID == id_iter & julian_day == jul_day_cutoff) %>% 
      pull(point_state)
    
    #report the cutoff date, ID, modified ending state, and real ending state in a tibble, which will be the output of the loop
    tibble(cutoff_date = jul_day_cutoff,
           ID = id_iter,
           variable_to_test = variable_to_test,
           end_modified = end_state_modified,
           end_true = end_state_true)
  },
  error = function(e){
    tibble(cutoff_date = jul_day_cutoff,
           ID = id_iter,
           variable_to_test = variable_to_test,
           end_modified = NA,
           end_true = NA)
  })
  return(result)
}
```


```{r}
cutoffs <- c("3/15/2020", "4/1/2020", "4/15/2020", "5/1/2020", "5/15/2020", "6/1/2020", "6/15/2020") %>% 
  mdy() %>% 
  yday()

#now iterating through individuals
set.seed(8)
id_subsample <- id_list %>% sample(size = 52)

parameters_to_run <- expand.grid(id_subsample, cutoffs) %>%
  mutate(rn = as.character(row_number()))
colnames(parameters_to_run) <- c("id_iter", "jul_day_cutoff", "rn")

parameters_to_run <- parameters_to_run %>% #if I omit the bird from the model, omit it from the accuracy evaluation
  filter(!(id_iter %in% c("RI-2019-21-2020", "RI-2019-28-2020", "RI-2019-29-2020", "VA-2018-03-2018", "FL-2021-01-2021")))
```

```{r}
parameters_to_run$valid <- parameters_to_run %>% 
  dplyr::select(id_iter, jul_day_cutoff) %>% 
  pmap(.f = function(id_iter, jul_day_cutoff){
    amwo_hmm_clipped <- amwo_hmm %>% #remove locations for the ID in question after the cutoff date
      filter(ID == id_iter & julian_day < jul_day_cutoff)
    
    observed_on_cutff <- amwo_hmm %>% #ensure that there's a sighting on the cutoff date
      filter(ID == id_iter & julian_day == jul_day_cutoff) %>% 
      nrow()
    
    logic_flag_1 <- nrow(amwo_hmm_clipped) > 1
    
    logic_flag_2 <- observed_on_cutff != 0
    
    if_else(logic_flag_1 & logic_flag_2, TRUE, FALSE) %>% 
      return()
  })

parameters_to_run <- parameters_to_run %>% 
  filter(valid == TRUE) %>% 
  dplyr::select(-valid)

#How many samples will this give me?
parameters_to_run %>%
  pull(id_iter) %>%
  unique() %>%
  length()

#build a sequence of models for variable selection
variable_iterations <- expand.grid(parameters_to_run$rn, names(Par0_m1)[c(3:7, 9)])

colnames(variable_iterations) <- c("rn", "variable_to_test")

variable_iterations <- variable_iterations %>% 
  mutate(rn = as.character(rn), variable_to_test = as.character(variable_to_test)) %>% 
  mutate(model_structure = map(.x = variable_to_test,
                              .f = function(i){
                                append(c("step", "angle", "step_500"), i) %>% 
                                  return()
                              }))
parameters_to_run <- parameters_to_run %>% 
  full_join(x = parameters_to_run, y = variable_iterations, by = c("rn"))

```

```{r}
plan(multisession, workers = 6)
tbl_list <- future_pmap(.l = parameters_to_run, 
                        .f = iterate_ids_male, 
                        .options = furrr_options(seed = 8),
                        .progress = TRUE)
plan(sequential)
```

```{r}
accuracy_results_male <- tbl_list %>% 
  do.call(rbind, .)

# Append results to previous runs
write.table(accuracy_results_male, "variable_importance_spring_male.csv",
            append = TRUE,
            sep = ",",
            col.names = TRUE,
            row.names = FALSE,
            quote = FALSE)
```
